Local PCA Shows How Population Structure Differs Along the Genome

### 2016-05-25 writeup/local_pca_paper.tex
added more refs

Realized kinship summarizes the shapes
of the genealogical that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
independent of the genomic region chosen.
This hope is not entirely justified: for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
and therefore patterns of kinship.
Indeed,
chromosome-scale variation in diversity and divergence has been observed in many species
\citep[e.g.][]{sattath2011pervasive};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citet{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor what the major axes of variation are.

@@

% To interpret the results of MDS, we combine known genome feature information for each species, such as the
% distribution of inversions, and heterochromatin and gene density along the genome. 
Each species showed distinct patterns, reflecting differences in their biology;
before presenting results,
it may help to have a prior idea of what factors are expected to affect population structure,
and how.
PCA summarizes patterns in kinship found in the genetic covariance matrix,
which is an average across locus-specific genealogies.
% However, different parts of genome have different genetic features. 
% First, each site of DNA may have different gene tree. 
% The covariance matrix of genotype matrix averages those gene trees. 
If individuals are closer in the genealogies of a given genomic region, 
they tends to be more close in the PCA maps for that region. 
% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
% Selective sweeps cause local recent ancestry or short trees.
% Background selection causes shallow ones.
Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
A genomic region with many targets for selection experiences
background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
which would tend to generally shorten genealogical trees in the region,
similar to a reduction in effective population size \citep{hudson1995deleterious}.
% Balancing selection causes deep trees.
On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
and population structure locally describes which individuals have which alleles
rather than geographical proximity.
Finally, 
since recombination is suppressed between opposite orientations of a chromosomal inversion
near its breakpoints,
genealogies in these regions separate samples carrying the two orientations of the inversion.
In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
as a way to identify inversions).
% if a chromosomal inversion is polymorphic in the sample, 
% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
% Other effects, like noise, introgression might also influence population structure.
We will look for strong variation in population structure shared across large regions of the genome.
Many of the effects listed above (such as single selective sweeps or inversions) 
are not expected to have similar effects on population structure in different regions of the genome
because of randomness in which samples end up in which group;
but if these coincide with a region of reduced recombination (such as an inversion),
these could drive major patterns of variation.
The more subtle effects of genome-wide linked selection could be shared across large regions
due to large-scale variation in gene density
(as by background selection or local adaptation at many genes).

@@

(sentence about how we found variation in pop structure)

With growing appreciation of the heterogeneous effects of selection across the genome,
especially the importance of adaptive introgression, hybrid speciation \citep{tomato,mosquito,introgression_review},
local adaptation \citep{local_adapt_review},
and inversion polymorphisms \citep{mosquito,kirkpatrick_inversions},
local PCA may prove to be a useful exploratory tool to discover important genomic features.
It is unclear whether the technique will be useful on reduced representation genotyping datasets 
due to marker density and issues with missing data --
our investigations with one such dataset were inconclusive --
but even low coverage, whole-genome sequence is very promising.

>>>>

Realized kinship summarizes the shapes
of the genealogical that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
independent of the genomic region chosen.
This hope is not entirely justified: for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
and therefore patterns of kinship.
Indeed,
chromosome-scale variation in diversity and divergence has been observed in many species
\citep[e.g.][]{langley2012genomic};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citet{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor what the major axes of variation are.

@@

% To interpret the results of MDS, we combine known genome feature information for each species, such as the
% distribution of inversions, and heterochromatin and gene density along the genome. 
Each species showed distinct patterns, reflecting differences in their biology;
before presenting results,
it may help to have a prior idea of what factors are expected to affect population structure,
and how.
PCA summarizes patterns in kinship found in the genetic covariance matrix,
which is an average across locus-specific genealogies.
% However, different parts of genome have different genetic features. 
% First, each site of DNA may have different gene tree. 
% The covariance matrix of genotype matrix averages those gene trees. 
If individuals are closer in the genealogies of a given genomic region, 
they tends to be more close in the PCA maps for that region. 
% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
% Selective sweeps cause local recent ancestry or short trees.
% Background selection causes shallow ones.
Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
A genomic region with many targets for selection experiences
background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
which would tend to generally shorten genealogical trees in the region,
similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
% Balancing selection causes deep trees.
On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
and population structure locally describes which individuals have which alleles
rather than geographical proximity.
Finally, 
since recombination is suppressed between opposite orientations of a chromosomal inversion
near its breakpoints,
genealogies in these regions separate samples carrying the two orientations of the inversion.
In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
as a way to identify inversions).
% if a chromosomal inversion is polymorphic in the sample, 
% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
% Other effects, like noise, introgression might also influence population structure.
We will look for strong variation in population structure shared across large regions of the genome.
Many of the effects listed above (such as single selective sweeps or inversions) 
are not expected to have similar effects on population structure in different regions of the genome
because of randomness in which samples end up in which group;
but if these coincide with a region of reduced recombination (such as an inversion),
these could drive major patterns of variation.
The more subtle effects of genome-wide linked selection could be shared across large regions
due to large-scale variation in gene density
(as by background selection or local adaptation at many genes).

@@

Our investigations have found a remarkable amount of variation in population structure across the genomes
of three diverse species,
revealing distinct biological processes driving this variation in each species.
More investigation, particularly on more species and datasets will help uncover which patterns are generalizable.

With growing appreciation of the heterogeneous effects of selection across the genome,
especially the importance of adaptive introgression, hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
local adaptation \citep{lenormand2002limits,wang2014isolation},
and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
local PCA may prove to be a useful exploratory tool to discover important genomic features.
It is unclear whether the technique will be useful on reduced representation genotyping datasets 
due to marker density and issues with missing data --
our investigations with one such dataset were inconclusive --
but even low coverage, whole-genome sequence is very promising.
